Sparse Function-space Representation \\ of Neural Networks

### 2023-05-14 paper/main.tex
typo
removed state from policy
add missing i
remove un needed i
added rasmussen chap

A common approach is to approximate the correlation structure of a distribution centred at the MAP estimate as done in the Laplace-GGN \citep{khan2019approximate, daxberger2021laplace, maddox2021fast}. The Laplace-GGN takes the MAP solution and approximates the Hessian of the loss function  
% $\ell(f_\weights(\mathbf{x}_{i}), \mathbf{y}_i)$ 
with the Generalized Gaussian-Newton approximation (GGN). 
Following the work of \citep{khan2019approximate}, we can view this approximation as building an approximate linear model of the neural network as $\hat{f}_\weights(\mathbf{x}) = \Jac{\weights_*}{\mathbf{x}} \, \weights$, where $\Jac{\weights}{\mathbf{x}} \coloneqq \nabla_\weights f(\mathbf{x})^\top \in \R^{C \times P}$ is the Jacobian at the MAP. Using the Hessian of the approximate model, we arrive at the Laplace-GGN posterior for $q(\vw)$.

In this paper, we are concerned with learning the posterior $q(\vf)$ directly in the function space, instead of $q(\vw)$. We choose to use a Gaussian process, an extension of the multivariate normal that characterizes a distribution over functions in terms of the first two moments, a mean $\mu(\cdot)$ and covariance  (or kernel) function $\kappa(\cdot,\cdot)$. We aim to derive these two functions based on the neural network to then build the Gaussian process posterior. Gaussian distributions remain tractable under learn transformations. Therefore, a linear function with respect to its parameters can be converted from the weight space to the function space (see Ch.~XXX in \cite{rasmussen2006gaussian} as follows: \todo{this part needs to be improved the linear function pops out of nowhere}
\begin{equation} \label{eq:weight_func}
g_\weights(\mathbf{x}_{i}) = \phi^\top\!(\vx) \, \vw \implies \mu(\vx) = 0 \quad \kappa(\mathbf{x}, \mathbf{x}') = \frac{1}{\delta} \phi(\vx)^\top \phi(\vx)
\end{equation}

@@

\section{Methods}
\paragraph{Linearization in the dual parameters}
As stated in \citet[Lemma~1]{csato2002sparse}, the posterior process mean and covariance functions of a GP prior, combined with a likelihood function and $\dataset$ can be stated as:
\begin{equation}  \label{eq:gp_pred}
 \myexpect_{p(f_i \mid\vy)}[f_i] = \vk_{\vx i}^\top \valpha \quad \text{and} \quad
\mathrm{Var}_{p(f\mid\vy)}[f_i] = \kappa_{ii} - \vk_{\vx i}^\top ( \MKxx + \diag(\vbeta)^{-1})^{-1} \vk_{\vx i},
\end{equation}
where $f_i = f_\weights(\vx_i)$, the $ij$\textsuperscript{th} entry of the matrix $\MKxx \in \R^{m \times m}$ is $\kappa(\vx_i,\vx_j)$, $\vk_{\vx i}$ denotes a vector where each $j$\textsuperscript{th} element is $\kappa(\vx_i, \vx_j)$, and $\kappa_{ii} = \kappa(\vx_i, \vx_i)$.  The dual parameters consist of the vector $\valpha \in \R^{m}$ and vector $\vbeta \in \R^{m}$, defined as:
\begin{equation}
\label{eq:dual_param}
\alpha_i \coloneqq \myexpect_{p(\vw \mid \vy)}[\nabla_{f_i}\log p(y_i \mid f_i)],  \quad
\beta_i \coloneqq - \myexpect_{p(\vw \mid \vy)}[\nabla^2_{f_i f_i}\log p(y_i \mid f_i)] ,
\end{equation}\todo{ More explination of the dual equation 6 the loss landscape link. Perbutation point linked directly to that here. }
%
The parameters $(\alpha_i, \beta_i)$ are the dual parameters that arise in the dual formulations used in, for example, support vector machines \citep{cortes1995support}, whose origins are found in work by \citet{kimeldorf1971some}. The above relations holds for generic likelihoods, and no approximations are involved as the expectation is under the exact posterior. The question is how do we form the dual variables without access to the exact posterior and different approximate inference techniques solve the question in different ways. Bound techniques such as \citet{jaakkola1997variational} attempt to bound the integral thus making the integral tractable, while Expecation Propogation (EP) as used in \citet{csato2002sparse} iteratively finds dual variables but sets them to log of the expectation of the approximate posterior. Meanwhile variational inference  \cite{khan2017conjugate, adam2021dual} find them using the variational approximation of the posterior to solve the integral.

@@

We consider stochastic environments with states \(\state \in \stateDomain \subseteq \R^{D_{\state}} \),
actions \(\action \in \actionDomain \subseteq \R^{D_{\action}}\) and transition dynamics
\(\transitionFn: \stateDomain \times \actionDomain \rightarrow \stateDomain \), such that
$\state_{t+1} = \transitionFn(\state_{t}, \action_{t}) + \noise_{t}$, where  $\noise_{t}$
is i.i.d. transition noise.
We consider the episodic setting where the system is reset to an initial state $\state_{0}$ at each episode and we
assume that there is a known reward function $r : \stateDomain \times \actionDomain \rightarrow \R$.
% Following the Markov decision process formulation \cite{bellmanMarkovianDecisionProcess1957a}, we
% denote the states \(\state \in \stateDomain \subseteq \R^{D_{\state}} \) and actions \(\action \in \actionDomain \subseteq \R^{D_{\action}}\)
% of the sytem, the reward function $r : \stateDomain \times \actionDomain \rightarrow \R$, and transition dynamics
% \(\transitionFn: \stateDomain \times \actionDomain \rightarrow \stateDomain \), such that
% $\state_{t+1} = \transitionFn(\state_{t}, \action_{t}) + \noise_{t}$ where  $\noise_{t}$
% is i.i.d. transition noise.
% $\mathbb{E}_{\noise_{0:\infty}} \big[ \sum_{t=0}^{\infty} \discount^{t} \rewardFn(\state_{t},\action_{t}) \big]$
% \begin{align} \label{eq-model-free-objective}
% \policy^{*} = \arg \max_{\policy \in \policyDomain} J(\transitionFn, \policy) = \arg \max_{\policy \in \policyDomain} \mathbb{E}_{\noise_{0:\infty}} \bigg[ \sum_{t=0}^{\infty} \discount^{t} \rewardFn(\state_{t},\action_{t}) \bigg]
% \quad \text{s.t. } \state_{t+1} = \transitionFn(\state_{t}, \action_{t}) + \noise_{t},
% \end{align}
The goal of RL is to find the policy \(\pi : \stateDomain \rightarrow \actionDomain\)
(from a set of policies $\Pi$) that maximises the sum of discounted rewards
in expectation over the transition noise,
\begin{align} \label{eq-model-free-objective}
J(\transitionFn, \policy) = \mathbb{E}_{\noise_{0:\infty}} \bigg[ \sum_{t=0}^{\infty} \discount^{t} \rewardFn(\state_{t},\action_{t}) \bigg]
\quad \text{s.t. } \state_{t+1} = \transitionFn(\state_{t}, \action_{t}) + \noise_{t},
\end{align}
where $\gamma \in [0, 1)$ is a discount factor.
In this work, we consider model-based RL where a model of the transition dynamics is learned \(f_{\mathbf{w}} \approx \transitionFn\) and then used by a planning algorithm.
A simple approach is to use the learned dynamic model $f_{\mathbf{w}^{*}}$ and maximise the objective in \cref{eq-model-free-objective},
\begin{align} \label{eq-greedy}
  \policy^{\text{Greedy}}(\state) &= \arg \max_{\pi \in \Pi} J(f_{\mathbf{w}^{*}}, \pi).
\end{align}
However, we can leverage the method in \cref{sec:methods} to obtain a function-space posterior over the learned dynamics $q_{\mathbf{u}}(\hat{\transitionFn} \mid \dataset)$,
where $\mathcal{D}$ represents the state transition data set \(\mathcal{D} = \{(s_{i},a_{i}), s_{i+1}\}_{i=0}^{N}\).
Importantly, the uncertainty represented by this posterior distribution can be used to balance the exploration-exploitation trade-off,
using approaches such as posterior sampling \cite{osbandWhyPosteriorSampling2017,osbandMoreEfficientReinforcement2013},
\begin{align} \label{eq-posterior-sampling}
  \policy^{\text{PS}}(\state) &= \arg \max_{\pi \in \Pi} J(\hat{f}, \pi)
\quad \text{s.t. } \tilde{\transitionFn} \sim q_{\mathbf{u}}(\hat{\transitionFn} \mid \dataset),
\end{align}
where a function $\tilde{\transitionFn}$ is sampled from the (approximate) posterior $q_{\mathbf{u}}(\hat{\transitionFn} \mid \dataset)$ and used to find a policy as
in \cref{eq-greedy}.
Intuitively, this strategy will explore where the model has high uncertainty, which in turn will reduce the model's uncertainty as data is collected and used to
train the model.
See \cref{sec-rl-appendix} for more details of how we use

@@

Our experiments seek to answer the following questions:
\begin{enumerate}
  \item \textbf{Predictions} How do predictions with our sparse function-space approximation compare to weight-space approximations? Does our method's ability to consider the full data set offer benefits over subset function-space methods?
  % \item Does our method's ability to consider the full data set offer benefits over subset function-space methods?
  \item \textbf{Function-space updates} How faster are our function-space updates relative to retraining from scratch? Do they improve predictive performance? Are they as good as retraining from scratch?
  \item \textbf{Uncertainty} How good are our uncertainty estimates? Can they be used in downstream settings like RL?
  \item \textbf{Representation} Is our sparse function-space representation useful for continual learning?
\end{enumerate}

>>>>

A common approach is to approximate the correlation structure of a distribution centred at the MAP estimate as done in the Laplace-GGN \citep{khan2019approximate, daxberger2021laplace, maddox2021fast}. The Laplace-GGN takes the MAP solution and approximates the Hessian of the loss function  
% $\ell(f_\weights(\mathbf{x}_{i}), \mathbf{y}_i)$ 
with the Generalized Gaussian-Newton approximation (GGN). 
Following the work of \citep{khan2019approximate}, we can view this approximation as building an approximate linear model of the neural network as $\hat{f}_\weights(\mathbf{x}) = \Jac{\weights_*}{\mathbf{x}} \, \weights$, where $\Jac{\weights}{\mathbf{x}} \coloneqq \nabla_\weights f_{\weights}(\mathbf{x})^\top \in \R^{C \times P}$ is the Jacobian at the MAP. Using the Hessian of the approximate model, we arrive at the Laplace-GGN posterior for $q(\vw)$.

In this paper, we are concerned with learning the posterior $q(\vf)$ directly in the function space, instead of $q(\vw)$. We choose to use a Gaussian process, an extension of the multivariate normal that characterizes a distribution over functions in terms of the first two moments, a mean $\mu(\cdot)$ and covariance  (or kernel) function $\kappa(\cdot,\cdot)$. We aim to derive these two functions based on the neural network to then build the Gaussian process posterior. Gaussian distributions remain tractable under learn transformations.
Therefore, a linear function with respect to its parameters can be converted from the weight space to the function space (see Ch.~2.1 in \cite{rasmussen2006gaussian}) as follows: \todo{this part needs to be improved the linear function pops out of nowhere}
\begin{equation} \label{eq:weight_func}
g_\weights(\mathbf{x}) = \phi^\top\!(\vx) \, \vw \implies \mu(\vx) = 0 \quad \kappa(\mathbf{x}, \mathbf{x}') = \frac{1}{\delta} \phi(\vx)^\top \phi(\vx)
\end{equation}

@@

\section{Methods}
\paragraph{Linearization in the dual parameters}
As stated in \citet[Lemma~1]{csato2002sparse}, the posterior process mean and covariance functions of a GP prior, combined with a likelihood function and $\dataset$ can be stated as:
\begin{equation}  \label{eq:gp_pred}
 \myexpect_{p(f_i \mid\vy)}[f_i] = \vk_{\vx i}^\top \valpha \quad \text{and} \quad
\mathrm{Var}_{p(f_{i}\mid\vy)}[f_i] = \kappa_{ii} - \vk_{\vx i}^\top ( \MKxx + \diag(\vbeta)^{-1})^{-1} \vk_{\vx i},
\end{equation}
where $f_i = f_\weights(\vx_i)$, the $ij$\textsuperscript{th} entry of the matrix $\MKxx \in \R^{m \times m}$ is $\kappa(\vx_i,\vx_j)$, $\vk_{\vx i}$ denotes a vector where each $j$\textsuperscript{th} element is $\kappa(\vx_i, \vx_j)$, and $\kappa_{ii} = \kappa(\vx_i, \vx_i)$.  The dual parameters consist of the vector $\valpha \in \R^{m}$ and vector $\vbeta \in \R^{m}$, defined as:
\begin{equation}
\label{eq:dual_param}
\alpha_i \coloneqq \myexpect_{p(\vw \mid \vy)}[\nabla_{f_i}\log p(y_i \mid f_i)],  \quad
\beta_i \coloneqq - \myexpect_{p(\vw \mid \vy)}[\nabla^2_{f_i f_i}\log p(y_i \mid f_i)] ,
\end{equation}\todo{ More explination of the dual equation 6 the loss landscape link. Perbutation point linked directly to that here. }
%
The parameters $(\alpha_i, \beta_i)$ are the dual parameters that arise in the dual formulations used in, for example, support vector machines \citep{cortes1995support}, whose origins are found in work by \citet{kimeldorf1971some}. The above relations holds for generic likelihoods, and no approximations are involved as the expectation is under the exact posterior. The question is how do we form the dual variables without access to the exact posterior and different approximate inference techniques solve the question in different ways. Bound techniques such as \citet{jaakkola1997variational} attempt to bound the integral thus making the integral tractable, while Expecation Propogation (EP) as used in \citet{csato2002sparse} iteratively finds dual variables but sets them to log of the expectation of the approximate posterior. Meanwhile variational inference  \cite{khan2017conjugate, adam2021dual} find them using the variational approximation of the posterior to solve the integral.

@@

We consider stochastic environments with states \(\state \in \stateDomain \subseteq \R^{D_{\state}} \),
actions \(\action \in \actionDomain \subseteq \R^{D_{\action}}\) and transition dynamics
\(\transitionFn: \stateDomain \times \actionDomain \rightarrow \stateDomain \), such that
$\state_{t+1} = \transitionFn(\state_{t}, \action_{t}) + \noise_{t}$, where  $\noise_{t}$
is i.i.d. transition noise.
We consider the episodic setting where the system is reset to an initial state $\state_{0}$ at each episode and we
assume that there is a known reward function $r : \stateDomain \times \actionDomain \rightarrow \R$.
% Following the Markov decision process formulation \cite{bellmanMarkovianDecisionProcess1957a}, we
% denote the states \(\state \in \stateDomain \subseteq \R^{D_{\state}} \) and actions \(\action \in \actionDomain \subseteq \R^{D_{\action}}\)
% of the sytem, the reward function $r : \stateDomain \times \actionDomain \rightarrow \R$, and transition dynamics
% \(\transitionFn: \stateDomain \times \actionDomain \rightarrow \stateDomain \), such that
% $\state_{t+1} = \transitionFn(\state_{t}, \action_{t}) + \noise_{t}$ where  $\noise_{t}$
% is i.i.d. transition noise.
% $\mathbb{E}_{\noise_{0:\infty}} \big[ \sum_{t=0}^{\infty} \discount^{t} \rewardFn(\state_{t},\action_{t}) \big]$
% \begin{align} \label{eq-model-free-objective}
% \policy^{*} = \arg \max_{\policy \in \policyDomain} J(\transitionFn, \policy) = \arg \max_{\policy \in \policyDomain} \mathbb{E}_{\noise_{0:\infty}} \bigg[ \sum_{t=0}^{\infty} \discount^{t} \rewardFn(\state_{t},\action_{t}) \bigg]
% \quad \text{s.t. } \state_{t+1} = \transitionFn(\state_{t}, \action_{t}) + \noise_{t},
% \end{align}
The goal of RL is to find the policy \(\pi : \stateDomain \rightarrow \actionDomain\)
(from a set of policies $\Pi$) that maximises the sum of discounted rewards
in expectation over the transition noise,
\begin{align} \label{eq-model-free-objective}
J(\transitionFn, \policy) = \mathbb{E}_{\noise_{0:\infty}} \bigg[ \sum_{t=0}^{\infty} \discount^{t} \rewardFn(\state_{t},\action_{t}) \bigg]
\quad \text{s.t. } \state_{t+1} = \transitionFn(\state_{t}, \action_{t}) + \noise_{t},
\end{align}
where $\gamma \in [0, 1)$ is a discount factor.
In this work, we consider model-based RL where a model of the transition dynamics is learned \(f_{\mathbf{w}} \approx \transitionFn\) and then used by a planning algorithm.
A simple approach is to use the learned dynamic model $f_{\mathbf{w}^{*}}$ and maximise the objective in \cref{eq-model-free-objective},
\begin{align} \label{eq-greedy}
  \policy^{\text{Greedy}} &= \arg \max_{\pi \in \Pi} J(f_{\mathbf{w}^{*}}, \pi).
\end{align}
However, we can leverage the method in \cref{sec:methods} to obtain a function-space posterior over the learned dynamics $q_{\mathbf{u}}(\hat{\transitionFn} \mid \dataset)$,
where $\mathcal{D}$ represents the state transition data set \(\mathcal{D} = \{(s_{i},a_{i}), s_{i+1}\}_{i=0}^{N}\).
Importantly, the uncertainty represented by this posterior distribution can be used to balance the exploration-exploitation trade-off,
using approaches such as posterior sampling \cite{osbandWhyPosteriorSampling2017,osbandMoreEfficientReinforcement2013},
\begin{align} \label{eq-posterior-sampling}
  \policy^{\text{PS}} &= \arg \max_{\pi \in \Pi} J(\hat{f}, \pi)
\quad \text{s.t. } \tilde{\transitionFn} \sim q_{\mathbf{u}}(\hat{\transitionFn} \mid \dataset),
\end{align}
where a function $\tilde{\transitionFn}$ is sampled from the (approximate) posterior $q_{\mathbf{u}}(\hat{\transitionFn} \mid \dataset)$ and used to find a policy as
in \cref{eq-greedy}.
Intuitively, this strategy will explore where the model has high uncertainty, which in turn will reduce the model's uncertainty as data is collected and used to
train the model.
See \cref{sec-rl-appendix} for more details of how we use

@@

Our experiments seek to answer the following questions:
\begin{enumerate}
  \item \textbf{Predictions} How do predictions with our sparse function-space approximation compare to weight-space approximations? Does our method's ability to consider the full data set offer benefits over subset function-space methods?
  % \item Does our method's ability to consider the full data set offer benefits over subset function-space methods?
  \item \textbf{Function-space updates} How fast are our function-space updates relative to retraining from scratch? Do they improve predictive performance? Are they as good as retraining from scratch?
  \item \textbf{Uncertainty} How good are our uncertainty estimates? Can they be used in downstream settings like RL?
  \item \textbf{Representation} Is our sparse function-space representation useful for continual learning?
\end{enumerate}
